### 2025-03-15 paper/quarto_ecml/paper.tex
Fix referencing inconsistencies and some final typos
Fin?

In the natural language processing (NLP) domain, CEs have also been used
to improve models through data augmentation. \cite{wu2021polyjuice2}
propose \emph{Polyjuice}, a general-purpose counterfactual generator for
language models. The authors empirically demonstrate that the
augmentation of training data with their method improves robustness in a
number of NLP tasks. \cite{balashankar2023improving} similarly use
\emph{Polyjuice} to augment NLP datasets through diverse counterfactuals
and show that classifier robustness improves by up to 20\%. Finally,
\cite{luu2023counterfactual} introduce Counterfactual Adversarial
Training (CAT), which also aims at improving generalization and
robustness of language models through a three-step procedure: the
authors identify training samples that are subject to high predictive
uncertainty, generate CEs for those samples, and then fine-tune the
language model on a dataset augmented with the CEs.

There have also been several attempts at formalizing the relationship
between counterfactual explanations and adversarial examples. Pointing
to clear similarities in how CEs and AEs are generated,
\cite{freiesleben2022intriguing} makes the case for jointly studying the
opaqueness and robustness problems in representation learning. Formally,
AEs can be seen as the subset of CEs for which misclassification is
achieved \cite{freiesleben2022intriguing}. Similarly,
\cite{pawelczyk2022exploring} show that CEs and AEs are equivalent under
certain conditions.

Two recent works are closely related to ours in that they use
counterfactuals during training with the explicit goal of affecting
certain properties of the post-hoc counterfactual explanations. Firstly,
\cite{ross2021learning} propose a way to train models that guarantee
individual recourse to some positive target class with high probability.
Their approach builds on adversarial training by explicitly inducing
susceptibility to targeted adversarial examples for the positive class.
Additionally, the proposed method allows for imposing a set of
actionability constraints ex-ante. For example, users can specify that
certain features are immutable. Secondly, \cite{guo2023counternet} are
the first to propose an end-to-end training pipeline that includes CEs
as part of the training procedure. In particular, they propose a
specific network architecture that includes a predictor and CE generator
network, where the parameters of the CE generator network are learnable.
Counterfactuals are generated during each training iteration and fed
back to the predictor network. In contrast to \cite{guo2023counternet},
we impose no restrictions on the ANN architecture at all.

@@

Improving the adversarial robustness of models is not the only path
towards aligning representations with plausible explanations. In a work
closely related to this one, \cite{altmeyer2024faithful} show that
explainability can be improved through model averaging and refined model
objectives. The authors propose a way to generate counterfactuals that
are maximally faithful to the model in that they are consistent with
what the model has learned about the underlying data. Formally, they
rely on tools from energy-based modelling \cite{teh2003energy} to
minimize the divergence between the distribution of counterfactuals and
the conditional posterior over inputs learned by the model. Their
proposed counterfactual explainer, \emph{ECCCo}, yields plausible
explanations if and only if the underlying model has learned
representations that align with them. The authors find that both deep
ensembles \cite{lakshminarayanan2016simple} and joint energy-based
models (JEMs) \cite{grathwohl2020your} tend to do well in this regard.

Once again it helps to look at these findings through the lens of
representation learning with high degrees of freedom. Deep ensembles are
approximate Bayesian model averages, which are most called for when
models are underspecified by the available data \cite{wilson2020case}.
Averaging across solutions mitigates the aforementioned risk of relying
on a single locally optimal representations that corresponds to
semantically meaningless explanations for the data. Previous work by
\cite{schut2021generating} similarly found that generating plausible
(``interpretable'') CEs is almost trivial for deep ensembles that have
also undergone adversarial training. The case for JEMs is even clearer:
they involve a hybrid objective that induces both high predictive
performance and generative capacity \cite{grathwohl2020your}. This is
closely related to the idea of aligning models with plausible
explanations and has inspired our CT objective.

@@

\begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
}\end{equation} Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
perturbation to any single feature is at most \(\varepsilon\). This is
closely aligned with \cite{szegedy2013intriguing} who define an
adversarial attack as an ``imperceptible non-random perturbation''.
Thus, we choose to work with a different distinction between CE and AE
than \cite{freiesleben2022intriguing} who consider misclassification as
the key distinguishing feature of AE. One of the key observations of
this work is that we can leverage CEs during training and get
adversarial examples essentially for free, which can be used to reap the
aforementioned benefits of adversarial training.

@@

In all cases the counterfactuals are valid but their quality differs. In
panel (a), they are not plausible: they do not comply with the
distribution of the factuals in \(y^+\) to the point where they form a
clearly distinguishable cluster. In panel (b), they are highly
plausible, meeting the first objective of Def. \ref{def-explainability}.
In panel (c), the CEs involve substantial reductions in \emph{debt} for
younger applicants. By comparison, counterfactual paths are shorter on
average in panel (d) where we have protected the immutable \emph{age} as
described in Section~\ref{sec-constraints}. Due to the classifier's
lower sensitivity to \emph{age}, recommendations with respect to
\emph{debt} are much more homogenous and do not unfairly punish younger
individuals. These counterfactuals are also plausible with respect to
the mutable feature. Thus, we consider the model in panel (d) as the
most explainable according to Def. \ref{def-explainability}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  To what extent does the counterfactual training objective (as defined
  in Equation~\ref{eq-obj}) induce models to learn plausible
  explanations?
\item
  To what extent does the CT objective produce more favorable
  algorithmic recourse outcomes in the presence of actionability
  constraints?
\item
  What are the effects of hyperparameter selection wrt. the CT
  objective?
\end{enumerate}

@@

Our key outcome of interest is how well models perform with respect to
explainability (Def. \ref{def-explainability}). To this end, we focus
primarily on the plausibility and cost of faithfully generated
counterfactuals at test time. To measure the cost of counterfactuals, we
follow the standard convention of using distances (\(\ell_1\)-norm)
between factuals and counterfactuals as a proxy. For plausibility, we
assess how similar counterfactuals are to observed samples in the target
domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the
distance-based metric used by \cite{altmeyer2024faithful},

@@

For all datasets except \emph{OL} and across all test settings, the
average distance of CEs from observed samples in the target class is
reduced, indicating improved plausibility. The magnitude of improvements
varies. For the simple synthetic datasets, distance reductions range
from around 20-40\% (\emph{LS}, \emph{Moon}) to almost 60\%
(\emph{Circ}). For the real-world tabular datasets, improvements tend to
be smaller but still substantial, with around 10-15\% for \emph{CH},
11-28\% for \emph{GMSC}, 7-8\% for \emph{Cred} and around 3\% for
\emph{Adult}. For our only vision dataset (\emph{MNIST}), distances are
reduced by up to 9\%. The results for our proposed divergence metric are
qualitatively similar, but generally even more pronounced: for the
\emph{Circ} dataset, implausibility is reduced by almost 94\% to
virtually zero as we verified by the absolute outcome. Improvements for
other datasets range from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For
\emph{OL} the reduction is negative, consistent with the distance-based
metric. \emph{MNIST} is the only dataset for which the two metrics
disagree.

We find that CT can substantially improve the quality of explanations
learned by models without sacrificing predictive performance. Test
accuracy for CT is virtually identical to the baseline for \emph{Adult},
\emph{Circ}, \emph{LS}, \emph{Moon} and \emph{OL}, and even slightly
improved for \emph{Cred}. Exceptions to this general pattern are
\emph{MNIST}, \emph{CH} and \emph{GMSC}, for which we observe reduction
in test accuracy of 2, 5 and 15 percentage points, respectively. In this
context, we also note that we have not optimized the models (very small
networks) for predictive performance at all.

@@

As our results indicate, counterfactual training produces models that
are more explainable. Nonetheless, it brings about three important
limitations.\\
\indent \emph{CT increases the training time of models.} CT can be more
time-consuming than conventional training regimes. While higher numbers
of CEs per iteration positively impact the quality of solutions, they
also increase the amount of computations. Relatively small grids with
270 settings can take almost four hours for more demanding datasets on a
high-performance computing cluster with 34 2GB CPUs.\footnote{See
  supplementary appendix for computational details.} Three factors
attenuate this effect. First, CT amortizes the cost of CEs for the
training samples. Second, we find that it can retain its value when used
as a ``fine-tuning'' technique for conventionally-trained models. Third,
it yields itself to parallel execution, which we have leveraged for our
own experiments.\\
\indent \emph{Immutable features may have proxies.} We propose an
approach to protect immutable features and thus increase the
actionability of the generated CEs. However, it requires that model
owners define the mutability constraints for (all) features considered
by the model. Even if all immutable features are protected, there may
exist proxies that are mutable (and hence should not be protected) but
preserve enough information about the principals to hinder the
protections. Delineating actionability is a major undecided challenge in
the AR literature \cite{venkatasubramanian2020philosophical} impacting
the capacity of CT to fulfill its intended goal.\\
\indent \emph{Interventions on features may impact fairness.} We provide
a tool that allows practitioners to modify the sensitivity of a model
with respect to certain features, which may have implication for the
fair and equitable treatment of decision subjects. As protecting a set
of features leads the model to assign higher relative importance to
unprotected features, model owners could misuse our solution by
enforcing explanations based on features that are more difficult to
modify by some (group of) individuals. For example, consider the Adult
dataset used in our experiments, where \emph{workclass} or
\emph{education} may be more difficult to change for underpriviledged
groups. When applied irresponsibly, CT could result in an unfairly
assigned burden of recourse \cite{sharma2020certifai}, threatening the
equality of opportunity in the system \cite{bell2024fairness}.
Nonetheless, these phenomena are not specific to CT.

>>>>

In the natural language processing (NLP) domain, CEs have also been used
to improve models through data augmentation. \cite{wu2021polyjuice2}
proposes \emph{Polyjuice}, a general-purpose counterfactual generator
for language models. The authors empirically demonstrate that the
augmentation of training data with their method improves robustness in a
number of NLP tasks. \cite{balashankar2023improving} similarly uses
\emph{Polyjuice} to augment NLP datasets through diverse counterfactuals
and show that classifier robustness improves by up to 20\%. Finally,
\cite{luu2023counterfactual} introduces Counterfactual Adversarial
Training (CAT) that also aims to improve generalization and robustness
of language models through a three-step procedure: the authors identify
training samples that are subject to high predictive uncertainty,
generate CEs for them, and fine-tune the language model on a dataset
augmented with the CEs.

There have also been several attempts at formalizing the relationship
between counterfactual explanations and adversarial examples. Pointing
to clear similarities in how CEs and AEs are generated,
\cite{freiesleben2022intriguing} makes the case for jointly studying the
opaqueness and robustness problems in representation learning. Formally,
AEs can be seen as the subset of CEs for which misclassification is
achieved \cite{freiesleben2022intriguing}. Similarly,
\cite{pawelczyk2022exploring} shows that CEs and AEs are equivalent
under certain conditions.

Two recent works are closely related to ours in that they use
counterfactuals during training with the explicit goal of affecting
certain properties of the post-hoc counterfactual explanations. Firstly,
\cite{ross2021learning} proposes a way to train models that guarantee
individual recourse to some positive target class with high probability.
Their approach builds on adversarial training by explicitly inducing
susceptibility to targeted adversarial examples for the positive class.
Additionally, the proposed method allows for imposing a set of
actionability constraints ex-ante. For example, users can specify that
certain features are immutable. Secondly, \cite{guo2023counternet} is
the first to propose an end-to-end training pipeline that includes CEs
as part of the training procedure. In particular, they propose a
specific network architecture that includes a predictor and CE generator
network, where the parameters of the CE generator network are learnable.
Counterfactuals are generated during each training iteration and fed
back to the predictor network. In contrast to \cite{guo2023counternet},
we impose no restrictions on the ANN architecture at all.

@@

Improving the adversarial robustness of models is not the only path
towards aligning representations with plausible explanations. In a work
closely related to this one, \cite{altmeyer2024faithful} shows that
explainability can be improved through model averaging and refined model
objectives. The authors propose a way to generate counterfactuals that
are maximally faithful to the model in that they are consistent with
what the model has learned about the underlying data. Formally, they
rely on tools from energy-based modelling \cite{teh2003energy} to
minimize the divergence between the distribution of counterfactuals and
the conditional posterior over inputs learned by the model. Their
proposed counterfactual explainer, \emph{ECCCo}, yields plausible
explanations if and only if the underlying model has learned
representations that align with them. The authors find that both deep
ensembles \cite{lakshminarayanan2016simple} and joint energy-based
models (JEMs) \cite{grathwohl2020your} tend to do well in this regard.

Once again it helps to look at these findings through the lens of
representation learning with high degrees of freedom. Deep ensembles are
approximate Bayesian model averages, which are most called for when
models are underspecified by the available data \cite{wilson2020case}.
Averaging across solutions mitigates the aforementioned risk of relying
on a single locally optimal representations that corresponds to
semantically meaningless explanations for the data. Previous work of
\cite{schut2021generating} similarly found that generating plausible
(``interpretable'') CEs is almost trivial for deep ensembles that have
also undergone adversarial training. The case for JEMs is even clearer:
they involve a hybrid objective that induces both high predictive
performance and generative capacity \cite{grathwohl2020your}. This is
closely related to the idea of aligning models with plausible
explanations and has inspired our CT objective.

@@

\begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
}\end{equation} Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
perturbation to any single feature is at most \(\varepsilon\). This is
closely aligned with \cite{szegedy2013intriguing} that defines an
adversarial attack as an ``imperceptible non-random perturbation''.
Thus, we choose to work with a different distinction between CE and AE
than \cite{freiesleben2022intriguing} that considers misclassification
as the key distinguishing feature of AE. One of the key observations of
this work is that we can leverage CEs during training and get
adversarial examples essentially for free, which can be used to reap the
aforementioned benefits of adversarial training.

@@

In all cases the counterfactuals are all valid but their quality
differs. In panel (a), they are not plausible: they do not comply with
the distribution of the factuals in \(y^+\) to the point where they form
a clearly distinguishable cluster. In panel (b), they are highly
plausible, meeting the first objective of Def. \ref{def-explainability}.
In panel (c), the CEs involve substantial reductions in \emph{debt} for
younger applicants. By comparison, counterfactual paths are shorter on
average in panel (d) where we have protected the immutable \emph{age} as
described in Section~\ref{sec-constraints}. Due to the classifier's
lower sensitivity to \emph{age}, recommendations with respect to
\emph{debt} are much more homogenous and do not unfairly punish younger
individuals. These counterfactuals are also plausible with respect to
the mutable feature. Thus, we consider the model in panel (d) as the
most explainable according to Def. \ref{def-explainability}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  To what extent does the counterfactual training objective as it is
  defined in Equation~\ref{eq-obj} induce models to learn plausible
  explanations?
\item
  To what extent does the CT objective produce more favorable
  algorithmic recourse outcomes in the presence of actionability
  constraints?
\item
  What are the effects of hyperparameter selection wrt. the CT
  objective?
\end{enumerate}

@@

Our key outcome of interest is how well models perform with respect to
explainability (Def. \ref{def-explainability}). To this end, we focus
primarily on the plausibility and cost of faithfully generated
counterfactuals at test time. To measure the cost of counterfactuals, we
follow the standard convention of using distances (\(\ell_1\)-norm)
between factuals and counterfactuals as a proxy. For plausibility, we
assess how similar counterfactuals are to observed samples in the target
domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the
distance-based metric used in \cite{altmeyer2024faithful},

@@

For all datasets except \emph{OL} and across all test settings, the
average distance of CEs from observed samples in the target class is
reduced, indicating improved plausibility. The magnitude of improvements
varies. For the simple synthetic datasets, distance reductions range
from around 20-40\% (\emph{LS}, \emph{Moon}) to almost 60\%
(\emph{Circ}). For the real-world tabular datasets, improvements tend to
be smaller but still substantial, with around 10-15\% for \emph{CH},
11-28\% for \emph{GMSC}, 7-8\% for \emph{Cred}, and around 3\% for
\emph{Adult}. For our only vision dataset (\emph{MNIST}), distances are
reduced by up to 9\%. The results for our proposed divergence metric are
qualitatively similar, but generally even more pronounced: for the
\emph{Circ} dataset, implausibility is reduced by almost 94\% to
virtually zero as we verified by the absolute outcome. Improvements for
other datasets range from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For
\emph{OL} the reduction is negative, consistent with the distance-based
metric. \emph{MNIST} is the only dataset for which the two metrics
disagree.

We find that CT can substantially improve the quality of explanations
learned by models without sacrificing predictive performance. Test
accuracy for CT is virtually identical to the baseline for \emph{Adult},
\emph{Circ}, \emph{LS}, \emph{Moon}, and \emph{OL}, and even slightly
improved for \emph{Cred}. Exceptions to this general pattern are
\emph{MNIST}, \emph{CH}, and \emph{GMSC}, for which we observe reduction
in test accuracy of 2, 5, and 15 percentage points respectively. In this
context, we also note that we have not optimized the models (very small
networks) for predictive performance at all.

@@

As our results indicate, counterfactual training produces models that
are more explainable. Nonetheless, it brings about three important
limitations.\\
\indent \emph{CT increases the training time of models.} CT can be more
time-consuming than conventional training regimes. While higher numbers
of CEs per iteration positively impact the quality of solutions, they
also increase the amount of computations. Relatively small grids with
270 settings can take almost four hours for more demanding datasets on a
high-performance computing cluster with 34 2GB CPUs.\footnote{See
  supplementary appendix for computational details.} Three factors
attenuate this effect. First, CT amortizes the cost of CEs for the
training samples. Second, we find that it can retain its value when used
as a ``fine-tuning'' technique for conventionally-trained models. Third,
it yields itself to parallel execution, which we have leveraged for our
own experiments.\\
\indent \emph{Immutable features may have proxies.} We propose an
approach to protect immutable features and thus increase the
actionability of the generated CEs. However, it requires that model
owners define the mutability constraints for (all) features considered
by the model. Even if all immutable features are protected, there may
exist proxies that are mutable (and hence should not be protected) but
preserve enough information about the principals to hinder the
protections. Delineating actionability is a major undecided challenge in
the AR literature (see, e.g.,
\cite{venkatasubramanian2020philosophical}) impacting the capacity of CT
to fulfill its intended goal.\\
\indent \emph{Interventions on features may impact fairness.} We provide
a tool that allows practitioners to modify the sensitivity of a model
with respect to certain features, which may have implication for the
fair and equitable treatment of decision subjects. As protecting a set
of features leads the model to assign higher relative importance to
unprotected features, model owners could misuse our solution by
enforcing explanations based on features that are more difficult to
modify by some (group of) individuals. For example, consider the Adult
dataset used in our experiments, where \emph{workclass} or
\emph{education} may be more difficult to change for underpriviledged
groups. When applied irresponsibly, CT could result in an unfairly
assigned burden of recourse \cite{sharma2020certifai}, threatening the
equality of opportunity in the system \cite{bell2024fairness}.
Nonetheless, these phenomena are not specific to CT.

### 2025-03-15 paper/sections/experiments.qmd
Fix referencing inconsistencies and some final typos
Fin?

1. To what extent does the counterfactual training objective (as defined in @eq-obj) induce models to learn plausible explanations?
2. To what extent does the CT objective produce more favorable algorithmic recourse outcomes in the presence of actionability constraints?
3. What are the effects of hyperparameter selection wrt. the CT objective?

@@

Our key outcome of interest is how well models perform with respect to explainability (Def. \ref{def-explainability}). To this end, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time. To measure the cost of counterfactuals, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy. For plausibility, we assess how similar counterfactuals are to observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$. We rely on the distance-based metric used by @altmeyer2024faithful,

@@

For all datasets except *OL* and across all test settings, the average distance of CEs from observed samples in the target class is reduced, indicating improved plausibility. The magnitude of improvements varies. For the simple synthetic datasets, distance reductions range from around 20-40% (*LS*, *Moon*) to almost 60% (*Circ*). For the real-world tabular datasets, improvements tend to be smaller but still substantial, with around 10-15% for *CH*, 11-28% for *GMSC*, 7-8% for *Cred* and around 3% for *Adult*. For our only vision dataset (*MNIST*), distances are reduced by up to 9%. The results for our proposed divergence metric are qualitatively similar, but generally even more pronounced: for the *Circ* dataset, implausibility is reduced by almost 94% to virtually zero as we verified by the absolute outcome. Improvements for other datasets range from 28% (*Moon*) to 78% (*GMSC*). For *OL* the reduction is negative, consistent with the distance-based metric. *MNIST* is the only dataset for which the two metrics disagree.

We find that CT can substantially improve the quality of explanations learned by models without sacrificing predictive performance. Test accuracy for CT is virtually identical to the baseline for *Adult*, *Circ*, *LS*, *Moon* and *OL*, and even slightly improved for *Cred*. Exceptions to this general pattern are *MNIST*, *CH* and *GMSC*, for which we observe reduction in test accuracy of 2, 5 and 15 percentage points, respectively. In this context, we also note that we have not optimized the models (very small networks) for predictive performance at all.

>>>>

1. To what extent does the counterfactual training objective as it is defined in @eq-obj induce models to learn plausible explanations?
2. To what extent does the CT objective produce more favorable algorithmic recourse outcomes in the presence of actionability constraints?
3. What are the effects of hyperparameter selection wrt. the CT objective?

@@

Our key outcome of interest is how well models perform with respect to explainability (Def. \ref{def-explainability}). To this end, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time. To measure the cost of counterfactuals, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy. For plausibility, we assess how similar counterfactuals are to observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$. We rely on the distance-based metric used in @altmeyer2024faithful,

@@

For all datasets except *OL* and across all test settings, the average distance of CEs from observed samples in the target class is reduced, indicating improved plausibility. The magnitude of improvements varies. For the simple synthetic datasets, distance reductions range from around 20-40% (*LS*, *Moon*) to almost 60% (*Circ*). For the real-world tabular datasets, improvements tend to be smaller but still substantial, with around 10-15% for *CH*, 11-28% for *GMSC*, 7-8% for *Cred*, and around 3% for *Adult*. For our only vision dataset (*MNIST*), distances are reduced by up to 9%. The results for our proposed divergence metric are qualitatively similar, but generally even more pronounced: for the *Circ* dataset, implausibility is reduced by almost 94% to virtually zero as we verified by the absolute outcome. Improvements for other datasets range from 28% (*Moon*) to 78% (*GMSC*). For *OL* the reduction is negative, consistent with the distance-based metric. *MNIST* is the only dataset for which the two metrics disagree.

We find that CT can substantially improve the quality of explanations learned by models without sacrificing predictive performance. Test accuracy for CT is virtually identical to the baseline for *Adult*, *Circ*, *LS*, *Moon*, and *OL*, and even slightly improved for *Cred*. Exceptions to this general pattern are *MNIST*, *CH*, and *GMSC*, for which we observe reduction in test accuracy of 2, 5, and 15 percentage points respectively. In this context, we also note that we have not optimized the models (very small networks) for predictive performance at all.

### 2025-03-15 paper/sections/related.qmd
Fix referencing inconsistencies and some final typos
Fin?

In the natural language processing (NLP) domain, CEs have also been used to improve models through data augmentation. @wu2021polyjuice2 propose *Polyjuice*, a general-purpose counterfactual generator for language models. The authors empirically demonstrate that the augmentation of training data with their method improves robustness in a number of NLP tasks. @balashankar2023improving similarly use *Polyjuice* to augment NLP datasets through diverse counterfactuals and show that classifier robustness improves by up to 20%. Finally, @luu2023counterfactual introduce Counterfactual Adversarial Training (CAT), which also aims at improving generalization and robustness of language models through a three-step procedure: the authors identify training samples that are subject to high predictive uncertainty, generate CEs for those samples, and then fine-tune the language model on a dataset augmented with the CEs.

There have also been several attempts at formalizing the relationship between counterfactual explanations and adversarial examples. Pointing to clear similarities in how CEs and AEs are generated, @freiesleben2022intriguing makes the case for jointly studying the opaqueness and robustness problems in representation learning. Formally, AEs can be seen as the subset of CEs for which misclassification is achieved [@freiesleben2022intriguing]. Similarly, @pawelczyk2022exploring show that CEs and AEs are equivalent under certain conditions.

Two recent works are closely related to ours in that they use counterfactuals during training with the explicit goal of affecting certain properties of the post-hoc counterfactual explanations. Firstly, @ross2021learning propose a way to train models that guarantee individual recourse to some positive target class with high probability. Their approach builds on adversarial training by explicitly inducing susceptibility to targeted adversarial examples for the positive class. Additionally, the proposed method allows for imposing a set of actionability constraints ex-ante. For example, users can specify that certain features are immutable. Secondly, @guo2023counternet are the first to propose an end-to-end training pipeline that includes CEs as part of the training procedure. In particular, they propose a specific network architecture that includes a predictor and CE generator network, where the parameters of the CE generator network are learnable. Counterfactuals are generated during each training iteration and fed back to the predictor network. In contrast to @guo2023counternet, we impose no restrictions on the ANN architecture at all.

@@

Improving the adversarial robustness of models is not the only path towards aligning representations with plausible explanations. In a work closely related to this one, @altmeyer2024faithful show that explainability can be improved through model averaging and refined model objectives. The authors propose a way to generate counterfactuals that are maximally faithful to the model in that they are consistent with what the model has learned about the underlying data. Formally, they rely on tools from energy-based modelling [@teh2003energy] to minimize the divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by the model. Their proposed counterfactual explainer, *ECCCo*, yields plausible explanations if and only if the underlying model has learned representations that align with them. The authors find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your] tend to do well in this regard.

Once again it helps to look at these findings through the lens of representation learning with high degrees of freedom. Deep ensembles are approximate Bayesian model averages, which are most called for when models are underspecified by the available data [@wilson2020case]. Averaging across solutions mitigates the aforementioned risk of relying on a single locally optimal representations that corresponds to semantically meaningless explanations for the data. Previous work by @schut2021generating similarly found that generating plausible ("interpretable") CEs is almost trivial for deep ensembles that have also undergone adversarial training. The case for JEMs is even clearer: they involve a hybrid objective that induces both high predictive performance and generative capacity [@grathwohl2020your]. This is closely related to the idea of aligning models with plausible explanations and has inspired our CT objective.

>>>>

In the natural language processing (NLP) domain, CEs have also been used to improve models through data augmentation. @wu2021polyjuice2 proposes *Polyjuice*, a general-purpose counterfactual generator for language models. The authors empirically demonstrate that the augmentation of training data with their method improves robustness in a number of NLP tasks. @balashankar2023improving similarly uses *Polyjuice* to augment NLP datasets through diverse counterfactuals and show that classifier robustness improves by up to 20%. Finally, @luu2023counterfactual introduces Counterfactual Adversarial Training (CAT) that also aims to improve generalization and robustness of language models through a three-step procedure: the authors identify training samples that are subject to high predictive uncertainty, generate CEs for them, and fine-tune the language model on a dataset augmented with the CEs.

There have also been several attempts at formalizing the relationship between counterfactual explanations and adversarial examples. Pointing to clear similarities in how CEs and AEs are generated, @freiesleben2022intriguing makes the case for jointly studying the opaqueness and robustness problems in representation learning. Formally, AEs can be seen as the subset of CEs for which misclassification is achieved [@freiesleben2022intriguing]. Similarly, @pawelczyk2022exploring shows that CEs and AEs are equivalent under certain conditions.

Two recent works are closely related to ours in that they use counterfactuals during training with the explicit goal of affecting certain properties of the post-hoc counterfactual explanations. Firstly, @ross2021learning proposes a way to train models that guarantee individual recourse to some positive target class with high probability. Their approach builds on adversarial training by explicitly inducing susceptibility to targeted adversarial examples for the positive class. Additionally, the proposed method allows for imposing a set of actionability constraints ex-ante. For example, users can specify that certain features are immutable. Secondly, @guo2023counternet is the first to propose an end-to-end training pipeline that includes CEs as part of the training procedure. In particular, they propose a specific network architecture that includes a predictor and CE generator network, where the parameters of the CE generator network are learnable. Counterfactuals are generated during each training iteration and fed back to the predictor network. In contrast to @guo2023counternet, we impose no restrictions on the ANN architecture at all.

@@

Improving the adversarial robustness of models is not the only path towards aligning representations with plausible explanations. In a work closely related to this one, @altmeyer2024faithful shows that explainability can be improved through model averaging and refined model objectives. The authors propose a way to generate counterfactuals that are maximally faithful to the model in that they are consistent with what the model has learned about the underlying data. Formally, they rely on tools from energy-based modelling [@teh2003energy] to minimize the divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by the model. Their proposed counterfactual explainer, *ECCCo*, yields plausible explanations if and only if the underlying model has learned representations that align with them. The authors find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your] tend to do well in this regard.

Once again it helps to look at these findings through the lens of representation learning with high degrees of freedom. Deep ensembles are approximate Bayesian model averages, which are most called for when models are underspecified by the available data [@wilson2020case]. Averaging across solutions mitigates the aforementioned risk of relying on a single locally optimal representations that corresponds to semantically meaningless explanations for the data. Previous work of @schut2021generating similarly found that generating plausible ("interpretable") CEs is almost trivial for deep ensembles that have also undergone adversarial training. The case for JEMs is even clearer: they involve a hybrid objective that induces both high predictive performance and generative capacity [@grathwohl2020your]. This is closely related to the idea of aligning models with plausible explanations and has inspired our CT objective.
